---
title: "Class 17: Analyzing sequencing data in the cloud"
author: "Linh Dang (PID: A16897764)"
format: pdf
toc: true
---

# Downstream analysis

Download `tximport()` in the **tximport** package to import Kallisto results into R.

```{r}
library(tximport)
```

```{r}
# setup the folder and filenames to read
folders <- dir(pattern="SRR21568*")
samples <- sub("_quant", "", folders)
files <- file.path(folders, "abundance.h5" )
names(files) <- samples
```

```{r}
txi.kallisto <- tximport(files, type = "kallisto", txOut = TRUE)
```

```{r}
head(txi.kallisto$counts)
```

How many transcripts we have for each sample?

```{r}
colSums(txi.kallisto$counts)
```

How many transcripts are detected in at least 1 sample?

```{r}
sum(rowSums(txi.kallisto$counts)>0)
```

Filter out transcripts with no reads and no changes over the sample:

```{r}
to.keep <- rowSums(txi.kallisto$counts) > 0
kset.nonzero <- txi.kallisto$counts[to.keep,]
keep2 <- apply(kset.nonzero,1,sd)>0
x <- kset.nonzero[keep2,]
```

# Principal Component Analysis

Perform a PCA of these samples, centering and scaling each transcript’s measured levels so that each feature contributes equally to the PCA:

```{r}
pca <- prcomp(t(x), scale=TRUE)
summary(pca)
```

Plot PC1 and PC2 to visualize samples:

```{r}
plot(pca$x[,1], pca$x[,2],
     col=c("blue","blue","red","red"),
     xlab="PC1", ylab="PC2", pch=16)
```

Do the same but for PC1 and PC3:

```{r}
plot(pca$x[,1], pca$x[,3],
     col=c("blue","blue","red","red"),
     xlab="PC1", ylab="PC3", pch=16)
```

Do the same but for PC2 and PC3:

```{r}
plot(pca$x[,2], pca$x[,3],
     col=c("blue","blue","red","red"),
     xlab="PC2", ylab="PC3", pch=16)
```

```{r}
# Make metadata object for the samples
colData <- data.frame(condition = factor(rep(c("control", "treatment"), each = 2)))
rownames(colData) <- colnames(txi.kallisto$counts)

# Make the data.frame for ggplot 
y <- as.data.frame(pca$x)
y$Condition <- as.factor(colData$condition)
```

Use `ggplot()` to plot PC1 and PC2:

```{r}
library(ggplot2)
library(ggrepel)
```

```{r}
ggplot(y) +
  aes(PC1, PC2, col=Condition) +
  geom_point() +
  geom_text_repel(label=rownames(y)) +
  theme_bw()
```

Plot PC1 and PC3:
```{r}
ggplot(y) +
  aes(PC1, PC3, col=Condition) +
  geom_point() +
  geom_text_repel(label=rownames(y)) +
  theme_bw()
```

Plot PC2 and PC3

```{r}
ggplot(y) +
  aes(PC2, PC3, col=Condition) +
  geom_point() +
  geom_text_repel(label=rownames(y)) +
  theme_bw()
```

Based on the PCA, we can clearly indicate that PC1 separate differences of the control samples from the treatment samples, while PC2 separate the differences between the two control samples (SRR2156848 and SRR2156849). 

# Differential Expression Analysis

```{r}
sampleTable <- data.frame(condition = factor(rep(c("control", "treatment"), each = 2)))
rownames(sampleTable) <- colnames(txi.kallisto$counts)
```

```{r, message=FALSE}
library(DESeq2)
```

```{r}
dds <- DESeqDataSetFromTximport(txi.kallisto,
                                sampleTable, 
                                ~condition)
dds <- DESeq(dds)
```

```{r}
res <- results(dds)
head(res)
```

