---
title: "Class 11: Comparative structure analysis"
author: "Linh Dang (PID:A16897764)"
format: pdf
---

Load up the packages we will need for analysis of protein structure sets.

```{r}
library(bio3d)
```

We will analyze the ADK starting with a string ADK database accession code: "1ake_A"

```{r}
id <- "1ake_A"
aa <- get.seq(id)
aa
```

Now we can search the PDB database to find all related enteries.

```{r}
blast <- blast.pdb(aa)
```

```{r}
attributes(blast)
```
```{r}
head(blast$hit.tbl)
```

Make a little summary figure of these results:

```{r}
hit <- plot(blast)
```

Our "top hits" i.e. the most similar enteries in the database are:

```{r}
hits <- NULL
hits$pdb.id <- c('1AKE_A','6S36_A','6RZE_A','3HPR_A','1E4V_A','5EJE_A','1E4Y_A','3X2S_A','6HAP_A','6HAM_A','4K46_A','3GMT_A','4PZL_A')
```


```{r}
# Download releated PDB files
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)
```

Align and superpose all these structures

```{r}
# Align releated PDBs
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")
```

Side-note:

```{r}
library(bio3dview)

#view.pdbs(pdbs)
```

This is better but still difficult to see what is similar and different in all these structures or indeed learn much about how this family works,

Let's try PCA:

```{r}
pc <- pca(pdbs)
plot(pc)
```

```{r}
plot(pc, pc.axes = 1:2)
```

```{r}
#view.pca(pc)
```

Write a PDB "trajectory" for mol-star

```{r}
mktrj(pc, file = "pca_results.pdb")
```




